Co-methylation review

### 2014-11-06 Rmarkdown/comethylation_review.Rmd
Completed modifications of a modification
Updated co-methylation review

## Chapter overview

Co-methylation is the dependence structure of DNA methylation data. This dependence structure can be viewed from different perspectives and on different scales, which has led to several disparate definitions of co-methylation. In this chapter I review these definitions and put them in the statistical framework of Chapter \ref{chap:stat_model}. I highlight various shortcomings of proposed methods, which led me to develop the methods and software described in Chapter \ref{chap:co-methylation}. I also review how methods to detect differential methylation have sought to account for and exploit co-methylation.

@@

Within-sample co-methylation is the spatial dependence of DNA methylation along the genome within a single sample. Many researchers have observed that there is a strong spatial dependence of DNA methylation along a genome, that is, methylation loci near to one another in the genome tend to be similarly methylated (e.g. \cite{Eckhardt:2006gh, Cokus:2008fc, Lister:2009hy, Lacey:2009by, Li:2010fb, Lyko:2010drm, Landan:2012kp}). Moreover, these studies found that the strength of this dependence itself depends on the distance between the loci and on the genomic context. The lowest level of within-sample co-methylation is the dependence structure of DNA methylation events occurring on the same DNA fragment, which I call _within-fragment co-methylation__. At a higher level is the _correlation of aggregate methylation levels_, such as the autocorrelation of $\beta$-values within a sample.

Between-sample co-methylation is the relationship between methylation levels at two or more loci across a set of samples[^concordance]. For example, \cite{Akulenko:2013fl} reported 187 pairs of genes whose methylation level was highly correlated (Pearson $|r| \geq 0.75$) across more than 300 breast cancer samples. Furthermore, measure of within-sample co-methylation can themselves be compared between samples, as is discussed in Chapter \ref{chap:co-methylation}.

@@

There have been several attempts to analyse co-methylation, most with the aim of better understanding the biological processes that lay down and regulate DNA methylation. An increased understanding of co-methylation could also lead to cheaper and more efficient assays of DNA methylation. For example, we could identify methylation loci whose methylation status is highly predictive of the methylation level of a surrounding region and design an assay to interrogate those highly predictive loci. This idea is analogous to the use of tag-SNPs on whole-genome genotyping microarrays (__CITE__).

@@

The analysis in \cite{Lister:2009hy} takes a different approach. It too uses within-sample aggregate methylation levels, but rather than estimating correlations of the methylation levels, instead looks for patterns in the spacing of "methylcytosines" along the genome.

Here I review this previous work. I have been forced to make my "best guess" about some of the methods since they are often not sufficiently described in the original publication.

### \cite{Eckhardt:2006gh}

\cite{Eckhardt:2006gh} is an important paper and one of the first to apply bisulfite-sequencing to a large number of regions and samples. Of particular relevance to a discussion of co-methylation, \cite{Eckhardt:2006gh} has been cited as evidence that "methylation levels are strongly spatially correlated" \cite{Hebestreit:2013ko} and that "proximal CpGs [have] similar methylation levels" \cite{Hansen:2011gu}. \cite{Eckhardt:2006gh} performed  PCR-amplicon bisulfite-sequencing of 2,524 amplicons on human chromosomes 6, 20 and 22.  PCR-amplicon bisulfite-sequencing is a labour-intensive assay and so it was a considerable effort to produce these data. Nonetheless, it is important to bear in mind that these data are nowhere near as comprehensive as data generated using modern-day whole-genome bisulfite-sequencing assays. Samples were provided by male and female donors across a wide variety of tissue types and ages, and were were generated to begin a catalogue of CpG methylation in various tissue types.

@@

In Figure 3a, the authors plot the "percentage identical methylation" against "distance in bp", which ranges from $0$ to $20,000$ bp (__TODO: Check permission on re-priting this figure__). The average amplicon length is reported as $411$ bp with a standard deviation of $71$ bp. Therefore, it is not possible that the "percentage identical methylation" was obtained from methylation events co-occuring on the same amplicon, at least for those pairs separated by more than approximately $600$ bp. Nowhere in the original paper is it clearly described what the "percentage identical methylation" is nor how it is computed. An email discussion with the senior author, Stephan Beck, was unable to clarify the definition.

My best guess as to the definition of "percentage identical methylation" is that it is the proportion of pairs of methylation calls at different CpGs from (generally) different amplicons where the methylation states are identical, that is, $Pr(z_{h, i} = z_{h', i'})$. I will refer to my best guess of "percentage identical methylation" as $PIM^{*}$, and now turn to some of its properties.

Let $0 \leq p_{h} \leq 1$ be the genome-wide average CpG methylation level for the sample from which $z_{i, h}$ is sampled, that is, $Pr(z_{h, i} = 1) = p_{h}$. For two distinct CpGs sequenced on two (potentially) distinct amplicons, $z_{h, i}$ and  $z_{h', i'}$, we want to compute the probability that the methylation states are identical, that is, $Pr(z_{h, i} = z_{h', i'})$. Not only may the two CpGs be from distinct amplicons, but the amplicons themselves may be from two distinct samples with different genome-wide average methylation levels. This most general setting is described by $z_{h, i} \eqd \text{Bernoulli}(p_{h})$ and $z_{h, i'} \eqd \text{Bernoulli}(p_{h'})$, with $cor(z_{h, i}, z_{h', i'}) = c$. In this setting we can compute the theoretical $PIM^{*}$ as follows:

\begin{align*}
PIM^{*} &= Pr(z_{h, i} = z_{h', i'}) \\
        &= Pr(z_{h, i} = 0, z_{h', i'} = 0) + Pr(z_{h, i} = 1, z_{h', i'} = 1) \\
        &= Pr(z_{h', i'} = 0 | z_{h, i} = 0) Pr(z_{h, i} = 0) +
           Pr(z_{h', i'} = 1 | z_{h, i} = 1) Pr(z_{h, i} = 1) \\
        &= \big [ (1 - p_{h'}) + c (1 - p_{h}) \big ] (1 - p_{h}) +
           \big [ (p_{h'} + c (1 - p_{h})) \big ] p_{h} \\
        &= (1 - p_{h}) (1 - p_{h'}) + p_{h} p_{h'} + 2 c p_{h} (1 - p_{h})
\end{align*}

This result follows by substituting $z_{h, i}$ and $p_{h}$ (resp. $z_{h', i'}$ and $p_{h'}$) for $X_{1}$ and $p_{1}$ (resp. $X_{2}$ and $p_{2}$) into the result derived in __TODO: Cite and link to appendix__.

@@

1. $PIM^{*}_{inid} = (1 - p_{h}) (1 - p_{h'}) + p_{h} p_{h'} \in [0, 1]$: Independent and non-identically distributed (_inid_), i.e., $z_{h, i} \eqd Bernoulli(p_{h})$ and $z_{h, i'} \eqd Bernoulli(p_{h'})$ and with $cor(z_{h, i}, z_{h', i'}) = c = 0$.
2. $PIM^{*}_{did} = (1 - p)^{2} + p^{2} + 2 c p ( 1 - p)$: Dependent and identically distributed (_did_), i.e., $z_{h, i}, z_{h', i'} \eqd Bernoulli(p)$ and $cor(z_{h, i}, z_{h', i'}) = c \neq 0$.
3. $PIM^{*}_{iid} = (1 - p)^{2} + p^{2}$: Independent and identically distributed (_iid_), i.e., $z_{h, i}, z_{h', i'} \eqd Bernoulli(p)$ and $cor(z_{h, i}, z_{h', i'}) = c = 0$.

@@

To return to the plot shown in Figure 3a; the "percentage identical methylation" asymptotes at $40 \%$, or $0.4$, which is less than the theoretical lower-bound of $PIM^{*}_{iid}$. If we believed that all samples had the same average CpG methylation levels, $p_{h}$, then we would conclude that the methylation states of two CpGs separated by a large distance are in fact __anti-correlated__ ($c < 0$) for CpGs separated by approximately $1,000$ bp. However, since we know that there is sample-to-sample variability in the average level of CpG methylation, the more likely explanation for this result is that the $p_{h}$ vary between samples and therefore $PIM^{*} < 0.5$ are possible even supposing $c > 0$.

@@

### \cite{Cokus:2008fc}\label{sec:Cokus}

@@

### \cite{Chodavarapu:2010iqa}

@@

### \cite{Li:2010fb}

@@

### \cite{Lyko:2010dr}

In their study of differential methylation in queen and worker honey bees (_Apis mellifera_), \cite{Lyko:2010dr} perform an autocorrelation analysis of CpG methylation (Supplementary Figure 7). They report that the "correlation of methylation status of neighboring CpGs increases sharply between 1 bp and 20 bp, then drops rapidly between 40 bp and 100 bp, and then slowly fades away". This is the first report of an organism in which the correlation of methylation levels first increases before decreasing (__TODO: Is this the "first and only" such report? Check my data any such increases__). __TODO: Is this the fastest decay of correlation in any species?__ However, I am concerned that the maximum autocorrelation is less $0.015$, which doesn't seem to make sense and is very close to zero. This may simply be due to a mislabelled or "normalised" value being plotted as the "correlation". (__TODO: I have emailed the senior author, Ryszard Maleszka, to try to resolve this concern__).

Additionally, \cite{Lyko:2010dr} identify a $3$ nucleotide periodicity in the autocorrelation of CpG methylation and provide evidence that this "reflect[s] a preferential methylation of CpGs occupying the first and second position of the arginine codons", which is coded for by the DNA triplets CGT, CGC, CGA, CGG, AGA, and AGG (__CITE__). They do not detect a $10$ nucleotide periodicity, which the authors claim is due to "a much lower density of modified CpGs in this species [_Apis mellifera_]" than in human or _Arabidopsis_ genomes.

Since the authors specifically describe their analysis as an "autocorrelation", I believe it uses all pairs of CpGs regardless of the number of intervening CpGs.

### \cite{Lacey:2013iy}

\cite{Lacey:2013iy} fit a Gaussian variogram model to the $\beta$-values at CpGs obtained via reduced representation bisulfite-sequencing (RRBS) of chromosome 11 from a normal myotube cell line (MTCTL2). The empirical and fitted variograms flatten out for CpGs separated by around $3000$ bp (Figure 1d) and so they conclude that the methylation level of CpGs "show a strong correlation for sites in close proximity, decaying to near independence at distances beyond 3000 bp".

@@

### \cite{Lister:2009hy}

@@

An $8-10$ nucleotide periodicity in the fitted cubic spline is observed in most, but not all, of these nine plots. This periodicity is very weak several contexts and not at all present for exonic CpGs. The weakness of the signal is partly attributable to the small sample sizes in the CHG and CHH contexts since there are very few "methylcytosines" for these contexts in these human samples. Furthermore, in all nine contexts the reported $8-10$ nucleotide periodicity is weaker than a $3$ nucleotide periodicity for "methylcytosines" at exonic CpGs; this $3$ nucleotide periodicity can clearly be observed in the raw counts, unlike the $8-10$ nucleotide periodicity, but is not commented on by the authors.

Based on these nine plots, \cite{Lister:2009hy} draw parallels to the $10$ periodicity in the autocorrelation of CHH $\beta$-values observed by \cite{Cokus:2008fc} and speculate that "_DNMT3A_  may be responsible for catalysing the methylation at non-CG sites". However, I suspect that the results in these nine plots are driven by periodicities in where cytosines are located in the genome rather than any periodicities in the actual methylation of these cytosines. That is not to say that there aren't periodicities in the methylation of cytosines, but that is not what is being measured in this analysis. To explain, these plots use the raw counts of "methylcytosines", which are not adjusted for the frequency of pairs of cytosines with a given $IPD$. If pairs of cytosines are observed more often at certain $IPD$s then even if methylation was totally independent of $IPD$ we would expect more pairs of "methylcytosines" at these same $IPD$s. Therefore, at a minimum, it is necessary to normalise the number of "methylcytosines" by the number of cytosines.

@@

### \cite{Landan:2012kp}

__TODO: Review co-methylation work in this paper__

@@

Because every sequencing read is produced from a single DNA fragment[^chimera], we can study within-fragment co-methylation by analysing the methylation patterns along individual sequencing reads. That is, by studying the methylation patterns at m-tuples. Furthermore, since the DNA methyltransferases act on individual DNA fragments (__TODO: check and cite__), these estimates of within-fragment co-methylation are on the same scale as the physical process that lays down and maintains DNA methylation.

@@

Long-read sequencing technologies, such as Pacific Biosciences (__CITE__), Roche 454 (__CITE__) and Oxford Nanopore (__CITE__), would overcome this limitation but the lesser throughput of these technologies and the reliability with which they can identify methylcytosines create new complications (__CITE__). An alternative solution, although completely hypothetical, is to use a technique that generates synthetic long reads, similar to those generated by Illumina's TruSeq Synthetic Long-Read technique[^moleculo](__CITE__).

@@

To summarise, current technologies mean that we are limited to studying within-fragment co-methylation at loci that are within regions of approximately $200$ nucleotides. There is less previous work studying within-fragment co-methylation than there is for correlations of aggregate methylation levels. Here I review this previously published work.

### \cite{Cokus:2008fc}

@@

The sequencing technology at the time of this publication produced very short reads, on average only $31$ nucleotides of usable sequence. For the results shown in Supplementary Figure 14 of \cite{Cokus:2008fc}, they first identify all reads containing multiple CHH loci. Then, for a pair of CHH loci in the same read, they estimate the probability that the second CHH in the pair is methylated given that the first CHH in the pair is methylated. This is estimated by simply counting the number of times this occurs and dividing by the number of pairs of CHH loci that occur within the same read[^cokus_clarification]. They estimate this probability separately for each distance between CHH loci and separately for each of the five _Arabidopsis_ chromosomes. \cite{Cokus:2008fc} then plot the average across the five chromosomes as a function of the number of nucleotides between the CHH loci. This plot includes pointwise $95\%$ confidence intervals, computed under an assumption of Normality, and a running average of these averages (). There is an approximate $10$ nucleotide periodicity evident in this plot, supporting the interpretation of the results observed in the autocorrelation of the $\beta$-values.

[^cokus_clarification]: Actually, it's not clear whether they consider all pairs of CHH loci or only the first pair of CHH loci in a read since a read may contain multiple pairs. The number of pairs of CHH loci that occur within the same read is not the same as the number of reads containing multiple CHH loci. Either method should give similar results since there will be few 31 nucleotide reads containing multiple pairs of CHH loci.

I consider these results to use a restricted definition of within-fragment "co-methylation". These results tells us nothing about the probability that both CHH loci are _unmethylated_ nor the joint probability distribution of methylation for two CHH loci on the same fragment. Specifically, \cite{Cokus:2008fc} only estimate $Pr(Z_{h, i} = 1)$ and $Pr(Z_{h, i'} = 1 | Z_{h, i} = 1)$, where $Z_{h, i}$ and $Z_{h, i'}$ are two CHH loci on the same fragment $(i < i')$. From these two quantities we can only estimate two of the four joint probabilities, $Pr(Z_{h, i} = x, Z_{h, i} = y)$, namely, $Pr(Z_{h, i} = 1, Z_{h, i'} = 1)$ and $Pr(Z_{h, i} = 1, Z_{h, i'} = 0)$. It does not allow estimation of $Pr(Z_{h, i} = 0, Z_{h, i'} = 0)$, nor of $Pr(Z_{h, i} = 0, Z_{h, i'} = 1)$, since these require estimates of $Pr(Z_{h, i'} = 0 | Z_{h, i} = 0)$. While it is trivial to additionally estimate $Pr(Z_{h, i'} = 0 | Z_{h, i} = 0)$, I argue in Section \ref{subsec:wf_cometh} that it is more useful and simpler to  estimate the odds ratio, $\frac{Pr(Z_{h, i} = 1, Z_{h, i'} = 1) Pr(Z_{h, i} = 0, Z_{h, i'} = 0)}{Pr(Z_{h, i} = 1, Z_{h, i'} = 0) Pr(Z_{h, i} = 0, Z_{h, i'} = 1)}$, rather than the conditional probabilities, $Pr(Z_{h, i'} = 1 | Z_{h, i} = 1)$ and $Pr(Z_{h, i'} = 0 | Z_{h, i} = 0)$.

@@

### \cite{Lacey:2009by}

@@

In summary, this analysis uses high resolution measurements from a moderate number of clones from a moderate number of samples for a miniscule fraction of the human genome. While the methods and models are quite interesting, the data are so focused that the results obtained from these analyses cannot represent a genome-wide picture of within-fragment co-methylation.

>>>>

\section{Introduction}

Co-methylation is the dependence structure of DNA methylation data. This dependence structure can be viewed from different perspectives and on different scales, which has led to several disparate definitions of co-methylation. In this chapter I first review these definitions and put them in the statistical framework of Chapter \ref{chap:stat_model}. I highlight various shortcomings of proposed methods, which led me to develop the methods and software described in  Chapter \ref{chap:co-methylation}. I also review how methods to detect differential methylation have sought to account for and exploit co-methylation.

@@

Within-sample co-methylation is the spatial dependence of DNA methylation along the genome within a single sample. Many researchers have observed that there is a strong spatial dependence of DNA methylation along a genome, that is, methylation loci near to one another in the genome tend to be similarly methylated (e.g. \cite{Eckhardt:2006gh, Cokus:2008fc, Lister:2009hy, Lacey:2009by, Li:2010fb, Lyko:2010drm, Landan:2012kp}). Moreover, these studies found that the strength of this dependence itself depends on the distance between the loci and on the genomic context. The fundamental level of within-sample co-methylation is the dependence structure of DNA methylation events occurring on the same DNA fragment, which I call _within-fragment co-methylation__. At a higher level is the _correlation of aggregate methylation levels_, such as the autocorrelation of $\beta$-values within a sample.

Between-sample co-methylation is the relationship between methylation levels at a pair of loci across a set of samples[^concordance]. For example, \cite{Akulenko:2013fl} reported 187 pairs of genes whose methylation level was highly correlated (Pearson $|r| \geq 0.75$) across more than 300 breast cancer samples. Also note, however, that measures of within-sample co-methylation can themselves be compared between samples, which is discussed at the end of this chapter.

@@

There have been several attempts to analyse co-methylation, most with the aim of better understanding the biological processes that lay down and regulate DNA methylation. An increased understanding of co-methylation could also lead to cheaper and more efficient assays of DNA methylation. For example, we could identify methylation loci whose methylation status is highly predictive of the methylation level of a surrounding region and design an assay to interrogate those highly predictive loci. This idea is analogous to the use of tag-SNPs on whole-genome genotyping microarrays.

@@

The analysis in \cite{Lister:2009hy} takes a different approach. While it also uses within-sample aggregate methylation levels, rather than estimating correlations of the methylation levels, the autnors instead looks for patterns in the spacing of "methylcytosines" along the genome.

In reviewing this previous work, I have been forced to make my "best guess" about some of the methods since they are often not sufficiently described in the original publication.

\subsection{\cite{Eckhardt:2006gh}}\label{sec:eckhardt}

\cite{Eckhardt:2006gh} is one of the first publications to apply bisulfite-sequencing to a large number of regions and samples. Of particular relevance to a discussion of co-methylation, \cite{Eckhardt:2006gh} has been cited as evidence that "methylation levels are strongly spatially correlated" \cite{Hebestreit:2013ko} and that "proximal CpGs [have] similar methylation levels" \cite{Hansen:2011gu}. \cite{Eckhardt:2006gh} performed  PCR-amplicon bisulfite-sequencing of 2,524 amplicons on human chromosomes 6, 20 and 22.  PCR-amplicon bisulfite-sequencing is a labour-intensive assay and so it was a considerable effort to produce these data. Nonetheless, it is important to bear in mind that these data are nowhere near as comprehensive as data generated using modern-day whole-genome bisulfite-sequencing assays. Samples were provided by male and female donors across a wide variety of tissue types and ages, and were were generated to begin a catalogue of CpG methylation in various tissue types.

@@

In Figure 3a, the authors plot the "percentage identical methylation" against "distance in bp", which ranges from $0$ to $20,000$ bp (__TODO: Check permission on re-printing this figure__). Nowhere in the original paper is it clearly described what the "percentage identical methylation" is nor how it is computed[^beck]. The average amplicon length is reported as $411$ bp with a standard deviation of $71$ bp. Therefore, it is impossible for the "percentage identical methylation" to have been obtained from methylation events co-occuring on the same amplicon, at least for those pairs separated by more than approximately $600$ bp.

[^beck]: An email discussion with the senior author, Stephan Beck, was unable to clarify the definition or method.

My best guess as to the definition of "percentage identical methylation" is that it is the proportion of pairs of identical methylation calls at different CpGs from (generally) different amplicons, that is, $Pr(Z_{h, i} = Z_{h', i'})$. I will refer to my best guess of "percentage identical methylation" as $PIM^{*}$, and now turn to some of its properties.

Let $0 \leq p_{h} \leq 1$ be the genome-wide average CpG methylation level for the sample from which $Z_{i, h}$ is sampled, that is, $Pr(Z_{h, i} = 1) = p_{h}$. For two distinct CpGs sequenced on two (potentially) distinct amplicons, $Z_{h, i}$ and  $Z_{h', i'}$, we want to compute the probability that the methylation states are identical, that is, $Pr(Z_{h, i} = Z_{h', i'})$. Not only may the two CpGs be from distinct amplicons, but in the original analysis the amplicons themselves may be from two distinct samples with different genome-wide average methylation levels. This most general setting is described by $Z_{h, i} \eqd \text{Bernoulli}(p_{h})$ and $Z_{h, i'} \eqd \text{Bernoulli}(p_{h'})$, with $cor(Z_{h, i}, Z_{h', i'}) = c$. In this setting we can compute the theoretical $PIM^{*}$ as follows:

\begin{align*}
PIM^{*} &= Pr(Z_{h, i} = Z_{h', i'}) \\
        &= Pr(Z_{h, i} = 0, Z_{h', i'} = 0) + Pr(Z_{h, i} = 1, Z_{h', i'} = 1) \\
        &= Pr(Z_{h', i'} = 0 | Z_{h, i} = 0) Pr(Z_{h, i} = 0) +
           Pr(Z_{h', i'} = 1 | Z_{h, i} = 1) Pr(Z_{h, i} = 1) \\
        &= \big [ (1 - p_{h'}) + c (1 - p_{h}) \big ] (1 - p_{h}) +
           \big [ (p_{h'} + c (1 - p_{h})) \big ] p_{h} \\
        &= (1 - p_{h}) (1 - p_{h'}) + p_{h} p_{h'} + 2 c p_{h} (1 - p_{h})
\end{align*}

This result is proved in __TODO: Cite and link to appendix__.

@@

1. $PIM^{*}_{inid} = (1 - p_{h}) (1 - p_{h'}) + p_{h} p_{h'} \in [0, 1]$: Independent and non-identically distributed (_inid_), i.e., $Z_{h, i} \eqd Bernoulli(p_{h})$ and $Z_{h, i'} \eqd Bernoulli(p_{h'})$ and with $cor(Z_{h, i}, Z_{h', i'}) = c = 0$.
2. $PIM^{*}_{did} = (1 - p)^{2} + p^{2} + 2 c p ( 1 - p)$: Dependent and identically distributed (_did_), i.e., $Z_{h, i}, Z_{h', i'} \eqd Bernoulli(p)$ and $cor(Z_{h, i}, Z_{h', i'}) = c \neq 0$.
3. $PIM^{*}_{iid} = (1 - p)^{2} + p^{2}$: Independent and identically distributed (_iid_), i.e., $Z_{h, i}, Z_{h', i'} \eqd Bernoulli(p)$ and $cor(Z_{h, i}, Z_{h', i'}) = c = 0$.

@@

To return to the plot shown in Figure 3a; the "percentage identical methylation" asymptotes at $40 \%$, or $0.4$, which is less than the theoretical lower-bound of $PIM^{*}_{iid}$. If we believed that all samples had the same average CpG methylation levels, $p_{h}$, then we would conclude that the methylation states of two CpGs separated by a large distance are in fact __anti-correlated__ ($c < 0$) for CpGs separated by approximately $1,000$ bp. However, since we know that there is sample-to-sample variability in the average level of CpG methylation, the more likely explanation for this result is that the $p_{h}$ vary between samples and therefore $PIM^{*} < 0.5$ are possible, even supposing $c > 0$.

@@

\subsection{\cite{Cokus:2008fc}}\label{sec:Cokus}

@@

\subsection{\cite{Chodavarapu:2010iqa}}{sec:chodavarapu}

@@

\subsection{\cite{Li:2010fb}}\label{sec:li}

@@

\subsection{\cite{Lyko:2010dr}}\label{sec:lyko}

In their study of differential methylation in queen and worker honey bees (_Apis mellifera_), \cite{Lyko:2010dr} perform an autocorrelation analysis of CpG methylation. They report that the "correlation of methylation status of neighbouring CpGs increases sharply between 1 bp and 20 bp, then drops rapidly between 40 bp and 100 bp, and then slowly fades away". This is the first report of an organism in which the correlation of methylation levels first increases before decreasing (__TODO: Is this the "first and only" such report? Check my data any such increases__). __TODO: Is this the fastest decay of correlation in any species?__

\cite{Lyko:2010dr} also identify a $3$ nucleotide periodicity in the autocorrelation of CpG methylation and provide evidence that this "reflect[s] a preferential methylation of CpGs occupying the first and second position of the arginine codons", which is coded for by the DNA triplets CGT, CGC, CGA, CGG, AGA, and AGG (__CITE__). They do not detect a $10$ nucleotide periodicity, which the authors claim is due to "a much lower density of modified CpGs in this species [_Apis mellifera_]" than in human or _Arabidopsis_ genomes.

The data supporting these results, Supplementary Figure 7 of the original publication, appears weak at first glance since the maximum value of the autocorrelation is less than $0.015$. I believe that this is due an error in the computation of the autocorrelation. Specifically, what I suspect has been computed is the autocorrelation of all positions in the genome, not just CpGs, where non-CpGs have been assigned a $\beta$-value of zero.

To demonstrate this, let $\beta^{*} = (\beta^{*}_{i'}, \ldots, \beta^{*}_{n'})$ be this "augmented" vector of $\beta$-values:

$$
\beta^{*}_{i'} = \left\{
  \begin{array}{ll}
    0 & \text{if } i' \text{ is not a CpG}\\
    \beta_{i} & \text{if } i' \text{ is a CpG}
  \end{array}
\right.
$$

Using the `FF` chromosome 22 data, I computed and plotted (__FIGURE__) the correlations of all pairs of CpG $\beta$-values as well as the autocorrelation of the augmented vector $\beta^{*}$. From these plots we see that the autocorrelation of $\beta^{*}$ is a scaled version of the correlations of all pairs of CpG $\beta$-values. Therefore, while the autocorrelations are incorrectly computed, conclusions based on the relative height of peaks should be approximately correct.

```{r}
# TODO: The above described analysis and plots

```

\subsection{\cite{Lacey:2013iy}}\label{sec:lacey_2013}

\cite{Lacey:2013iy} fit a Gaussian variogram model to the $\beta$-values at CpGs obtained via reduced representation bisulfite-sequencing (RRBS) of chromosome 11 from a normal myotube cell line (MTCTL2). The empirical and fitted variograms asymptote for CpGs separated by around $3000$ bp (Figure 1d) and so they conclude that the methylation level of CpGs "show a strong correlation for sites in close proximity, decaying to near independence at distances beyond 3000 bp".

@@

\subsection{\cite{Lister:2009hy}}\label{sec:lister_2009}

@@

An $8-10$ nucleotide periodicity in the fitted cubic spline is observed in most, but not all, of these nine plots. This periodicity is very weak in several contexts and is absent for exonic CpGs. The weakness of the signal is partly attributable to the small sample sizes in the CHG and CHH contexts, since there are very few "methylcytosines" for these contexts in these human samples. Furthermore, in all nine contexts, the reported $8-10$ nucleotide periodicity is weaker than a $3$ nucleotide periodicity for "methylcytosines" at exonic CpGs; this $3$ nucleotide periodicity can clearly be observed in the raw counts, unlike the $8-10$ nucleotide periodicity, but is not commented on by the authors.

Based on these nine plots, \cite{Lister:2009hy} draw parallels to the $10$ periodicity in the autocorrelation of CHH $\beta$-values observed by \cite{Cokus:2008fc} and speculate that "_DNMT3A_  may be responsible for catalysing the methylation at non-CG sites". However, I suspect that the results in these nine plots are in fact driven by periodicities in where cytosines are located in the genome rather than any periodicities in the actual methylation of these cytosines. That is not to say that there aren't periodicities in the methylation of cytosines, but that is not what is being measured in this analysis. To explain, these plots use the raw counts of "methylcytosines", which are not adjusted for the frequency of pairs of cytosines with a given $IPD$. If pairs of cytosines are observed more often at certain $IPD$s then even if methylation was totally independent of $IPD$ we would expect more pairs of "methylcytosines" at these same $IPD$s. Therefore, at a minimum, it is necessary to normalise the number of "methylcytosines" by the number of cytosines.

@@

@@

Because every sequencing read is produced from a single DNA fragment[^chimera], we can study within-fragment co-methylation by analysing the methylation patterns along individual sequencing reads, that is, by studying the methylation patterns at m-tuples. Furthermore, since the DNA methyltransferases act on individual DNA fragments (__TODO: check and cite__), these estimates of within-fragment co-methylation are on the same scale as the physical process that lays down and maintains DNA methylation.

@@

Long-read sequencing technologies, such as Pacific Biosciences (__CITE__), Roche 454 (__CITE__) and Oxford Nanopore (__CITE__), would overcome this limitation but the lower throughput of these technologies, and the reliability with which they can identify methylcytosines, create new complications (__CITE__). An alternative solution, although completely hypothetical, is to use a technique that generates synthetic long reads, similar to those generated by Illumina's TruSeq Synthetic Long-Read technique[^moleculo](__CITE__).

@@

To summarise, current technologies mean that we are limited to studying within-fragment co-methylation at loci that are within regions of approximately $200$ nucleotides. There is less previous work studying within-fragment co-methylation than there is for correlations of aggregate methylation levels.

\subsection{\cite{Cokus:2008fc}}\label{sec:cokus}

@@

The sequencing technology at the time of this publication produced very short reads, on average only $31$ nucleotides of usable sequence. For the within-fragment co-methylation results (Supplementary Figure 14 of \cite{Cokus:2008fc}) they first identify all reads containing multiple CHH loci. Then, for a pair of CHH loci in the same read, they estimate the probability that the second CHH in the pair is methylated given that the first CHH in the pair is methylated. This is estimated by simply counting the number of times this occurs and dividing by the number of pairs of CHH loci that occur within the same read[^cokus_clarification]. They estimate this probability separately for each distance between CHH loci and separately for each of the five _Arabidopsis_ chromosomes. \cite{Cokus:2008fc} then plot the average across the five chromosomes as a function of the number of nucleotides between the CHH loci. This plot includes pointwise $95\%$ confidence intervals, computed under an assumption of Normality, and a running average of these averages. There is an approximate $10$ nucleotide periodicity evident in this plot, supporting the interpretation of the results observed in the autocorrelation of the $\beta$-values.

[^cokus_clarification]: Actually, it's not clear whether they consider all pairs of CHH loci or only the first pair of CHH loci in a read, since a read may contain multiple pairs. The number of pairs of CHH loci that occur within the same read is not the same as the number of reads containing multiple CHH loci. Either method should give similar results since there will be few 31 nucleotide reads containing multiple pairs of CHH loci.

I consider these results to use a restricted definition of within-fragment "co-methylation". To explain, these results tell us nothing about the probability that both CHH loci are _unmethylated_ nor the joint probability distribution of methylation for two CHH loci on the same fragment, but only about the probability that both CHH loci are methylated. Specifically, \cite{Cokus:2008fc} only estimate $Pr(Z_{h, i} = 1)$ and $Pr(Z_{h, i'} = 1 | Z_{h, i} = 1)$, where $Z_{h, i}$ and $Z_{h, i'}$ are two CHH loci on the same fragment $(i < i')$.

From these two quantities we can only estimate two of the four joint probabilities, $Pr(Z_{h, i} = x, Z_{h, i} = y)$, namely, $Pr(Z_{h, i} = 1, Z_{h, i'} = 1)$ and $Pr(Z_{h, i} = 1, Z_{h, i'} = 0)$. It does not allow estimation of $Pr(Z_{h, i} = 0, Z_{h, i'} = 0)$, nor of $Pr(Z_{h, i} = 0, Z_{h, i'} = 1)$, since these require estimates of $Pr(Z_{h, i'} = 0 | Z_{h, i} = 0)$. While it is trivial to additionally estimate $Pr(Z_{h, i'} = 0 | Z_{h, i} = 0)$, I argue in Section \ref{subsec:wf_cometh} that it is more useful and simpler to estimate the odds ratio, $\frac{Pr(Z_{h, i} = 1, Z_{h, i'} = 1) Pr(Z_{h, i} = 0, Z_{h, i'} = 0)}{Pr(Z_{h, i} = 1, Z_{h, i'} = 0) Pr(Z_{h, i} = 0, Z_{h, i'} = 1)}$, rather than the conditional probabilities, $Pr(Z_{h, i'} = 1 | Z_{h, i} = 1)$ and $Pr(Z_{h, i'} = 0 | Z_{h, i} = 0)$.

@@

\subsection{\cite{Lacey:2009by}}\label{sec:lacey_2009}

@@

In summary, this analysis uses high resolution measurements from a moderate number of clones from a moderate number of samples for a minuscule fraction of the human genome. While the methods and models are quite interesting, the data are so limited that the results obtained from these analyses cannot represent a genome-wide picture of within-fragment co-methylation.

\subsection{\cite{Landan:2012kp}}\label{sec:landan}

\cite{Landan:2012kp} is notable for its use of the within-read information available from bisulfite-sequencing data.  The authors explore the within-sample heterogeneity of methylation patterns at 4-tuples and 6-tuples of CpGs. They use ultradeep (median coverage of each CpG in the target $> 10,000 \times$) bisulfite-sequencing of a 45 cancer-related CpG islands and study how methylation dynamics at this regions evolves over time. They not only use this information to estimate co-methylation but also to estimate the rate of _epipolymorphisms_ and _epialleles_.

As part of their analysis they compute "methylation linkage diagrams". For each pair of CpGs in their targeted regions, they create a $2 \times 2$ contingency table. The general form of the table is shown below in Table \ref{tab:landan}. They then compute the Pearson correlation of each table[^cor_equivalence], which estimates the correlation of methylation events at that pair of CpGs. Importantly, these are within-read measurements from a single sample and so these are estimates of within-read, within-sample co-methylation.

[^cor_equivalence]: Since this is a correlation of two binary variables, it is equivalent to computing the _phi_ coefficient of each table (__TODO: Find a better source than Wikipedia and my own numerical calculations. The algebraic proof isn't hard, just tedious.__).

\begin{table}[h]
\centering
\begin{tabular}{@{}ccccc@{}}
\cmidrule(l){3-5}
 &  & \multicolumn{3}{c}{Second CpG} \\ \cmidrule(l){3-5}
 &  & m & u & Total \\ \cmidrule(l){2-5}
\multirow{3}{*}{First CpG} & m & $n_{m, m}$ & $n_{m, u}$ &  \\
 & u & $n_{u, m}$ & $n_{u, u}$ &  \\
 & Total &  &  & n \\ \cmidrule(l){2-5}
\end{tabular}
\caption{Summary of read counts at a CpG 2-tuple.}
\label{tab:landan}
\end{table}
